\begin{abstract}
    % Existing blockchain testing methodologies face significant limitations when evaluating time-controlled consensus protocols like Proof-of-Stake (PoS). 
    % Testnet and simulation are mainstream testing methods. 
    % Conventional testnets bind protocol time to wall-clock execution, creating a \emph{time lock} that severely restricts testing efficiency and reproducibility. While simulation bypasses the time-lock via abstracted logic and discrete-event scheduling, simulation sacrifices its fidelity and cannot be used to detect implementation-level bugs. 
        
    % We identify this bottleneck and propose a best-of-both solution: \emph{Event-Driven Testnet}, a methodology that treats time as a controllable input by decoupling logical protocol time from wall-clock execution, while running real clients. EDT requires a network simulator as the substrate and replaces the simulator's virtual node with a lightweight virtualization layer that binds the virtual-node scheduling interface to the execution of real, unmodified client binaries. 
    % On the basis of Distributed Virtual Network(DVN) and Shadow, we implement this event-driven testnet on Ethereum PoS and demonstrate substantial improvements in efficiency (up to $40\times$ for single-node testnet) without sacrificing fidelity, and experimental flexibility and reproducibility.
    % This enables faster, reproducible security evaluation, like attack replay, fuzzing at specific time intervals, and deterministic debugging. 
    
Existing methodologies for testing blockchain consensus protocols struggle to simultaneously provide fidelity, efficiency, and reproducibility for \emph{strongly time-controlled} protocols such as Proof-of-Stake (PoS). Conventional testnets bind logical protocol time to wall-clock execution, creating a \emph{time lock} that wastes most testing time waiting for slot/epoch boundaries. Simulation avoids this bottleneck through abstracted logic and discrete-event scheduling, but sacrifices implementation fidelity and therefore cannot reliably expose bugs in real clients.
    
We identify this coupling between protocol time and execution time as the core bottleneck and present a best-of-both \emph{time-decoupled testnet} methodology that breaks the time lock while still running real, unmodified client software. Our design runs real clients atop an event-driven execution substrate: a lightweight virtualization layer binds each virtual host's scheduling interface directly to the execution of real client processes under a unified simulation clock. This decouples logical protocol time from wall-clock time, makes protocol time explicitly controllable, and preserves protocol semantics without modifying protocol parameters.
We instantiate this approach for Ethereum PoS on top of Distributed Virtual Network (DVN) and Shadow. Experiments show that the resulting time-decoupled testnet preserves key consensus behaviors, achieves up to 40$\times$ wall-clock speedup over conventional testnets, and increases consensus core-logic execution density by orders of magnitude. Under fixed seeds and configurations, it also yields deterministic executions that support reproducible replay and practical distributed debugging of time-critical consensus behavior.
\end{abstract}